---
title: "Introduction to R"
author: "Joe Rudolf"
date: "5/22/2018"
output: html_document
---

# What is R?

R is a statistical programming language.  Using R you can load, analyze, and visualize data.  R has the same functionality you are used to in other spreadsheet applications (e.g. Microsoft Excel) but offers greater functionality and flexibility.  R also provides an environment in which we can conduct reproducible data analysis.  R affords us the capabilities to document our data analysis and save it in a way that we can revisit our analysis at a later date, update our approach when inspiration strikes, and share our analysis with friends. 

# Installing R (and some other stuff to make R work)

## 1.  R

R is the programming language that we will be using to conduct our data analysis today.

To install R, go to the [Comprehensive R Archive Network (CRAN)](https://cloud.r-project.org]).  Select the version of R for your operating system (Windows, Mac OS, Linux)

## 2.  RStudio

R studio is the development environment in which we will do our R programming.

Download and install [RStudio](https://www.rstudio.com/products/rstudio/download/#download).  Select the installer for your Operating System.

When you launch RStudio you will be presented with with a series of windows.

On the left you will see the *console*.  The console is where we will enter and execute our R code.  

The lower right window contains the *output*.  This is where the results of our code (e.g. our plots) will display.  

The window in the upper right corner of the screen contains our *history*.  Here we can see the historical record of the code we have executed in the console. 

## 3.  The Tidyverse

To enhance the functionality of R, we need to install additional packages.  A package is a collection of functions that extend the capabilities of the base R programming language.

In this course we will be primarily using the tidyverse package.  The tidyverse includes functions for reading data into the R environment, cleaning and manipulating data, and plotting our results.

Installing the tidyverse package is as easy as entering the following line of code in the RStudio console (the box on the left hand side of RStudio) and pressing the return (enter) key.

```{r}
install.packages("tidyverse", dependencies = TRUE)
```

# Configuring the RStudio Environment

## Setting the Working Directory

Let's create a folder on our desktop called "R_intro" in which to save our work today.  Now let's establish our R_intro folder as our working directory.

A working directory is where R will read our data from and save our work to.  Let's set our working directory to to the R_intro folder in RStudio (Session -> Set Working Directory -> Choose Directory).

We can check to see that we have set the correct working directory by running the following code in the console.

```{r}
getwd()
```

## Loading the tidyverse

We have previously installed the tidyverse package but to use the package in our analysis we need to load it during the session we intend to use the functions in.

Let's load the tidyverse with the following code.

```{r}
library(tidyverse)
```

# Basics of coding

In its most basic form R is a calculator

```{r}
2 + 3 + 2
```

R supports a number of functions (e.g. the absolute value function) that can extend our calculator. We pass arguments to the function (e.g. data or other parameters) and the function will give us back a result.

For example:

```{r}
abs(-77)
```
There are functions to support nearly all of the data analysis you may want to do.  And remember you can always add more functions by installing packages that other users have developed and published.  You can even write your own functions, though this won't be necessary for any of the analysis we perform in this introductory course.

We can also create variables (objects) and use them in our calculations.

To do this we use the following syntax (grammar):  

variable_name <- contents of our variable

For example:

```{r}
# Establish the variable
sequence_one_to_twenty <- seq(1, 20)

# Check the contents of the variable
sequence_one_to_twenty

# Calculate the median of the sequence
median(sequence_one_to_twenty)

# Calculate the max of the sequence
max(sequence_one_to_twenty)

```
Remember to give your variables descriptive names so you can easily remember the data they contain.  It is also a good practice to separate words in your variable names with underscores instead of spaces.

# Learning to Learn to Code

Learning to write code (including the functions available to you and the appropriate syntax to make your code work) can be frustrating!  Learning a new language takes time so be patient with yourself.  We all go through this process and your intution improves with your experience.

Inevitably, you will at times get stuck (especially when you are first starting).  However, there are some great resources to get you un-stuck and part of learning to code is learning how to leverage the wealth of knowledge of others.

## Help Command

If you know (or have an idea of a function) you would like to use, you can use the help command to find out more. Just type the "?" before the function of interest.


```{r}

?abs()

```

## The Internet

Chances are, if you are facing a data anlysis dilemma, many others have faced the exact same (or at least very similar) challenge.  Head to your favorite search engine.

Just ask your question (making sure to include "R" in there somewhere) and you will likely find what you are looking for in the first few results

"how to mutate column in r"

Chances are one of your first hits will be [Stack Overflow](https://stackoverflow.com/).  Stack Overflow is a great resource for developers to ask and answer questions.  If you have a question, it has likely been answered at least once on Stack Overflow.

## Cheat Sheets

The folks who make RStudio offer a number of really great [cheat sheets](https://www.rstudio.com/resources/cheatsheets/) on their website that you can download/print and keep next to your desk as quick references.

## A Good Book

Even in the 21st century, it never hurts to have a good book by your side.

The content in this course is oriented around the tidyverse.  If you would like to continue your studies leveraging the tidyverse, a great companion is R for Data Science.  This book is available freely in a digital format [here](http://r4ds.had.co.nz/) and print versions are available as well.

## Taking a Break

Sometimes you won't be able to brute force your way through a problem and your other resources (help functions, the internet, and your reference books) will seemingly let you down.  This can be very frustrating!  Don't despair, walk away for a bit.  Oftentimes it just takes a fresh set of eyes to reframe your question or rethink you approach, and what seemed impossible, suddenly becomes obvious.


# Sequencing functions

When you are working with a data set, you often need to manipulate it multiple times in a defined sequence of events. Let's start with a non-sensical example that can help illustrate the issue (adapted from the [tidyverse style guide](http://style.tidyverse.org/pipes.html)).

Let's say we want to apply the functions hop, scoop, and bop to the foo_foo data frame, in that order. One way to approach that is to start with the data, apply the function, and write the output back into the original data frame.

```{r, eval = FALSE}
# one way to represent a hop, scoop, and a bop, without pipes
foo_foo <- hop(foo_foo, through = forest)
foo_foo <- scoop(foo_foo, up = field_mice)
foo_foo <- bop(foo_foo, on = head)
```

R allows you to nest functions within one another, but this can get horribly confusing because following a specific sequence of operations requires you to start from the inside of the expression and expand outwards.

```{r, eval = FALSE}
# another way to represent the same sequence with less code but in a less readable way
foo_foo <- bop(scoop(hop(foo_foo, through = forest), up = field_mice), on = head)
```

You want to try and avoid doing things this way because the sequence of operations is so non-intuitive.

Explicitly showing the functions sequentially by line is helpful for readability but it does require some unnecessary typing to keep repeating the name of the data set. R allows you to "pipe" a data frame from one function to another using this funny looking operator: `%>%`. This can cut down on unnecessary code but also preserves the nice formatting that makes it obvious what functions are applied in what order.

```{r, eval = FALSE}
# a hop, scoop, and a bop with the almight pipes
foo_foo %>%
  hop(through = forest) %>%
  scoop(up = field_mouse) %>%
  bop(on = head)
```

Pipes are not compatible with all functions but should work with all of the tidyverse package functions (the magrittr package that defines the pipe is included in the tidyverse). In general, functions expect data as the primary argument and you can think of the pipe as feeding the data to the function. From the perspective of coding style, the most useful suggestion for using pipes is arguably to write the code so that each function is on its own line. The tidyverse style guide [section on pipes](http://style.tidyverse.org/pipes.html) is pretty helpful.

# Creating reproducible data analyis

Entering code in the console is a great way to develop and iterate your analysis.  However, it is easy to lose track of the pieces of code we have executed in the console. To truly make our code reproducible, we need to place it in a container that we can edit, save, and execute sequentially.  R Markdown files (extension .RMD) provide an authoring environment for us to do just that!

## Creating an R Markdown File

Let's create a simple R Markdown file together.

Go to the menu bar at the top of the screen (File -> New File -> R Markdown).

Let's get oriented to the R Markdown environment.

1. Header information (author, title, date, output format) is a the top of the document surrounded by "---".

2. R code is surrounded by "```" and will execute when the document is compiled using the knit button.

3. Commentary can be added to the document and formatted with simple formatting options (# for a title heading)

Copy and paste the following code into your R Markdown file.

```{r}

#First let's load the tidyverse

library(tidyverse)

#Load a data set of cars into a dataframe
cars_data <- as_tibble(mtcars)

#Display table
cars_data

#Fiter the list to cars with fuel economy >= 15 mpg and horsepower >=100
mpg_hp <-filter(mtcars, mpg >= 15, hp >=100)

#Create a scatter plot of mpg vs horsepower 
ggplot(data = mpg_hp) + geom_point(mapping = aes(x = mpg, y = hp))
```

Now let's save our R Markdown file to our working directory (File -> Save As) as "Our First R Markdown File.RMD"

## Running our code

To run our code we can use the Knit button at the top of the R Markdown window or copy/paste selections of our code into the console window and run it there.

What if we wanted to look at mpg >= 10 and re-run the analysis?  Let's change this parameter in our code and knit the file again.

R Markdown is a wonderful way to document our analyis and ensure its reproducibility.  We will plan to use R Markdown throughout the course and recommend that you incorporate R Markdown documentation as your preffered authoring tool.

## Sharing our code

Documents can be knitted as html, PDF, or word files.  These documents can be shared with your collaborators so others can see your results.  Additionally you can share the R Markdown file itself (along with any data files used for your analysis) so that others can run (and tweak) your analysis. 






