---
title: "Method Validation in R"
author: "Daniel Herman"
date: "06/07/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(readxl)
library(janitor)
```

# Overview

In this section we will walk through exampls of how to use R to analyze standard method validation experiments. We will use real data acquired as part of a method transition for immunoassays (unconjugated estriol, alpha fetoprotein, and inhibin-A) used for prenatal assessment of Trisomy 21, Trisomy 18, and open neural tube defect risk assessment.


# Load data

Relatively "clean" data has already been organized in separate tabs in an excel document. Let's first load one of these data sets. To do this we will use the `read_excel` function. Let's first take a look at this function. Click the green arrow to execute this code chunk:

```{r explore_function}
?read_excel
```

We see that there are three different functions grouped together, a `read_excel` function that is a wrapper for two more specific functions `read_xls` and `read_xlsx`. Since our Excel document is in .xlsx format, we can either use the general `read_excel` or `read_xlsx` function.

```{r load_data_1}
dat <- read_excel(path="data/Maternal Serum AFP HCG uE3 DxI raw data.deaccessioned.xlsx", 
                sheet="MS AFP")
```


Let's next take a look at the data

```{r}
dat
```

