---
title: "Exploratory Data Analysis:  Plotting Time Data and Joining Data Sets"
author: "Joe Rudolf"
date: "7/17/2018"
output: html_document
---

  
```{r setup}
knitr::opts_chunk$set(echo = TRUE, comment = NA)

library(tidyverse)
library(readxl)
library(janitor)
library(lubridate)

```

# Exploratory Data Analysis:  Plotting Time Data and Joining Data Sets

In this section, we will build on the previous section of exploring a data set.  We will introduce the concept of working with time data including calculating time differences.  This is very useful when working with clinical laboratory data (think turnaround times).  At the end of the lesson we will also discuss the basics of joining separate data sets together using 'join' functions.

Let's get started by loading the orders_data_set (which we used in the last lesson).  We will use the `read_excel` function and couple that to the `clean_names` function to tidy up those column names.

```{r}

orders <- read_excel("data/orders_data_set.xlsx") %>%
clean_names()

```

Now let's use `glimpse` to take a peak at our data.

```{r}

glimpse(orders)

```

So far we have worked with data types that include numbers and characters.  Take a look at order_time, result_time, and review_time in our glimpse.  They have a different data type <dttm>.  This stands for date time.  When we loaded the orders data R recognized this data as having the format of a date and time.  Having our data classified as date times gives us additional functionality to work with those data elements in interesting ways.  We can isolate components of a date and time, or perform date math (add and subtract dates from one another).

We can select just the date component of a date time:
  
```{r}

date_of_order <- date(orders$order_time)

print('Date of order:')

head(date_of_order)

```

Or the week component:
  
```{r}

week_of_order <- week(orders$order_time)

print('Week of order:')

head(week_of_order)

```

Here, the number "33" represents the 33rd week of the year.


**Exercise 1:** 

Can you isolate the day of the week from the order?  Hint: use the `wday` function.

```{r, eval=FALSE}

```

Here, the number "1" represents a Sunday.

For more interesting date time functions, check out the R Studio [Dates and Times Cheat Sheet](https://www.rstudio.com/resources/cheatsheets/).

**End exercise**

## Plotting Time Points for Data Exploration

Let's use the `hour` function to isolate the hour component of our order and then use `ggplot` to visualize the data.  This will give us a window into the time of day that ordering providers are placing their orders.

```{r}

order_hour <- hour(orders$order_time)

ggplot(data = orders) +
  geom_histogram(mapping = aes(x = order_hour), bins = 24)

```

Our data visualization shows that most orders are placed in the mid-morning and mid-afternoon with a break for lunch.  Also, as we might expect there aren't a lot of orders placed in the middle of the night.


**Exercise 2:**  

Can you perform a similar analysis and visualization to look at the result_time data using the `hour` function and `ggplot`?
 
```{r, eval=FALSE}

```

It looks like the lab gets busy issuing results around 5 in the morning and that the evening rush begins to tail off around 9-10 at night.

**End exercise**

## Exploring Time Intervals and Plotting the Data

So far in this lesson, we have looked at single time points.  It's also very useful to calculate and plot time differences.  Let's illustrate this concept by calculating the difference between the order and result times (turnaround time)

Not all tests that are ordered are completed.  We can look at a `summary` of the order status get a sense for this.

```{r}

summary(as.factor(orders$order_status_c_descr))

```


Results that aren't completed won't have a result time so we should `filter` our data first to only "Completed results".  That will allow us to work only with test results that have both an order time and result time.

```{r}

completed_orders <- orders %>%
  filter(order_status_c_descr == "Completed")

```

Now that we are working with completed orders only, we can add a column to our data frame and populate it with the difference between the order and result times.  We will use the `mutate` function to create the column and calculate the time difference.  We will also use the `filter` function to limit the window we are looking at to 24 hours (60 seconds x 60 minutes x 24 hours).  We are limiting our time window because the data has a long tail (think sendouts) that we will exclude so the plot is easier to read.  Then we can use `ggplot` to plot the data.


```{r}

order_to_result_delta <- completed_orders %>%
  mutate(order_time_to_result_time = result_time - order_time) %>%
  filter(order_time_to_result_time > 0 & order_time_to_result_time < 60*60*24)


ggplot(data = order_to_result_delta) + 
  geom_histogram(mapping = aes(x = order_time_to_result_time), binwidth = 60*60)

```


**Exercise 3:**  

What if we wanted to look at result time to review time instead?  Can you leverage the same approach as above to calculate and plot the difference between review time and the result time?  This will give us a look at the time between when results are available and when those results are reviewed by the ordering provider.

```{r, eval=FALSE}

```

**End exercise**

Note that the default date difference calculation is in seconds.  We can make some changes to our code to look a this data in other time components (such as hours). The `difftime` function allows us to specify a unit, for example hours in the example below.  We also have to make some changes to our `filter` logic and `ggplot` binwidth criteria to reflect that we are working with hours now and not seconds.

```{r}

order_to_result_delta <- completed_orders %>%
  mutate(order_time_to_result_time = difftime(result_time, order_time, units = "hour")) %>%
  filter(order_time_to_result_time > 0 & order_time_to_result_time < 24)

ggplot(data = order_to_result_delta) + 
  geom_histogram(mapping = aes(x = order_time_to_result_time), binwidth = 1)

```

## Counting Tests Per Patient

We will return to date time calculations in a bit but first let's quickly revisit tabulating data.  `Count` is a useful function for tabulation.  Say we wanted to know the number of completed tests per patient in our data set.  Since our data set has one test per patient per line, we can count the frequency of a patient's record number (MRN) as measure of the number of orders per patient.

```{r}

orders_count <- completed_orders %>%
  count(patient_id)

ggplot(data = orders_count) +
  geom_histogram(mapping = aes(x = n), binwidth = 1)

```

The vast majority of patient's received 10 or less tests during the time window that this data set covers.  It looks like there are a few patients that received as many as 50 tests.

**Exercise 4:**  

What if we wanted to know only number of CBCs per patient?  Can you use the `filter` function to limit the data CBC orders before counting?

```{r, eval=FALSE}

```

**End exercise**

## Date Differential by Patient and Test

Now that we have refreshed our tabulation skills, let's revist working with dates.  Sometimes it is useful to look at the interval between orders for a given test.  Let's look at the time interval between CBCs for individual patients in our data set.

To measure the difference in time between CBCs for a given patient we need to first make a combination of all of a patient's CBCs with all of the patient's other CBCs.  The `join` function allows us to join a data set to itself or another data set using a shared key, an attribute (or column) shared between the two data sets.  The nuances of joins exceed the scope of an introductory coding course, but we provide two examples below as templates for your future analysis.  If you would like to read more on joins, the "R for Data Science" book has a nice chapter on [Relational Data](http://r4ds.had.co.nz/relational-data.html).

Let's create the combination of patient CBCs, calculate the difference between CBC times, and display the output.


```{r}

cbc_orders <- filter(completed_orders, proc_code == "CBC")

cbc1 = select(cbc_orders, patient_id, result_time)
cbc2 = select(cbc_orders, patient_id, result_time)

cbc_join <- full_join(cbc1, cbc2, by= "patient_id")

cbc_join <- cbc_join %>%
  mutate(cbc_time_diff = difftime(result_time.x, result_time.y, units = "days")) %>%
  filter(cbc_time_diff > 0)

cbc_join <- cbc_join %>%
  group_by(patient_id, result_time.x) %>%
  summarize(minimum = min(cbc_time_diff))

cbc_join

```

This gives us the patient's record number, the result time of the CBC, and the amount of time that has elapsed since the patient's previous CBC.

Now let's plot the intervals using our old friend 'ggplot'.

```{r}

ggplot(data = cbc_join) + 
  geom_histogram(mapping = aes(x = minimum), binwidth = 5)

```

We can see that intervals between CBCs range from 1 to nearly 100 days in this data set, but most repeat CBCs are performed before ~45 days have elapsed.

## Bringing different data sets together:

The `join` function can also help us to bring two different data sets together for analysis using a shared key variable.

Let's load a second data set called order_details.  This data set includes information about the ordering route and whether the order originiated from a preference list.  This data set is a Comma Separated Variable (CSV) file so we need to use the `read_csv` function instead of our usual `read_excel` function.

```{r}

orders_details <- read_csv("data/order_details.csv")

head(orders_details)

```

We can join the order details data to our completed orders data using the order_id as the shared key for linking the data sets.

```{r}

orders_details <- read_csv("data/order_details.csv")

joined_data <- completed_orders %>%
  left_join(orders_details, by = "order_id") %>%
  count(pref_list_type)

joined_data

```

We can see that most completed orders originate from clinic prefernce lists, followed by individual provider preference lists, and then from searching the test menu in the electronic medical record (the "None" category)

What if we just wanted to look at preference list types for CBC?  We need to use the `filter` function and the proc_code of "CBC".

```{r}

orders_details <- read_csv("data/order_details.csv")

CBC <- completed_orders %>%
  filter(proc_code == "CBC") %>%
  left_join(orders_details, by = "order_id") %>%
  count(pref_list_type)

CBC

```

**Exercise 5:**  

What if we just wanted to look at Factor V Leiden.  Hint: proc_code = "F5DNA")

```{r, eval=FALSE}

```

**End exercise**