---
title: "Style Sheet Draft"
author: "Joe Rudolf"
date: "6/27/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Overview
A quick sentence describing what we will learn in this section

# 1 Heading

Description of learning and code to follow

```{r}

# Comment individual lines of code that may need additional explanation (help unpack how the functions function)


```

**Exercise:** Description of exercise for student to complete followed by blank r chunk (in version we distribute before class, completed for distrubtion afterwards)

```{r}
#Write your code here:


```



## Subheading



# 2 Heading


# 3 Heading


# Summary (Words of Wisdom, Recommendation for Further Study, Functions to check out)


---

Other thoughts:

Provide enough explanation that the document can stand on its own and attendees can revisit it later.

Write code so that it will knit successfully from front to back.

Leave Echo on so that attendees can see what R is doing?

Bold terms so they stand out **




